Open source symbolic multibody dynamics software

### 2012-02-10 PetersonGedeHubbard2012.tex
Fixed error in defintion of C1, C2, constrained state equations.
Made acceleration constraints distinct from dynamic equations.

The complete configuration of the system is assumed to be described by
generalized coordinates $\bm{q}\in\bm{R}^n$, its velocity completely described
by generalized speeds $\bm{u}\in\bm{R}^o$.  It is assumed that there are $l$
holonomic constraints ($f_h : \mathbf{R}^n \times \mathbf{R} \to
\mathbf{R}^l$), and $m$ nonholonomic constraints ($f_{nh} : \mathbf{R}^n \times
\mathbf{R}^o \times \mathbf{R} \to \mathbf{R}^l$). Two functions $\bm{f}_0 :
\mathbf{R}^n \times \mathbf{R}^n \times \mathbf{R} \to \mathbf{R}^n$ and
$\bm{f}_1 : \mathbf{R}^n \times \mathbf{R}^o \times \mathbf{R} \to
\mathbf{R}^n$ relate the coordinate time derivatives to the generalized speeds,
and two more functions $\bm{f}_2 : \mathbf{R}^n \times \mathbf{R}^o \times
\mathbf{R} \to \mathbf{R}^o$ and $\bm{f}_3 : \mathbf{R}^n
\times \mathbf{R}^n \times \mathbf{R}^o \times \mathbf{R}^p \times \mathbf{R}
\to \mathbf{R}^o$ relate the time derivatives of the speeds $\bm{\dot{u}}$ to
the externally applied inputs $\bm{r}\in\bm{R}^q$, coordinates $\bm{q}$ and
their time derivatives $\bm{\dot{q}}$,  speeds $\bm{u}$, and time $t$. These
assumptions can be summarized by the vector equations
\begin{align}
  \label{eq:holonomic}
  \bm{f}_{h}(\bm{q}, t) &= \bm{0}\\
  \label{eq:nonholonomic}
  \bm{f}_{nh}(\bm{q}, \bm{u}, t) &= \bm{0}\\
  \label{eq:kindiffs}
  \bm{f}_{0}(\bm{q}, \bm{\dot{q}}, t)
    + \bm{f}_{1}(\bm{q}, \bm{u}, t) &= \bm{0} \\
  \label{eq:dyndiffs}
  \bm{f}_{2}(\bm{q}, \bm{\dot{u}}, t)
    + \bm{f}_{3}(\bm{q}, \bm{\dot{q}}, \bm{u}, \bm{r}, t) & = \bm{0}
\end{align}
where equations (\ref{eq:holonomic}) and (\ref{eq:nonholonomic}) are the
holonomic and nonholonomic constraints (and possibly time differentiated holonomic
constraints), respectively. Equations (\ref{eq:kindiffs}) are the kinematic differential equations
(linear in both $\bm{\dot{q}}$ and $\bm{u}$ and in the trivial case,
$\bm{\dot{q}} = \bm{u}$, so $f_0 = -f_1$).  Equations (\ref{eq:dyndiffs}) are
the dynamic differential equations
and the time differentiated version of equation \ref{eq:nonholonomic}.
Equation (\ref{eq:dyndiffs}) will necessarily by linear in $\bm{\dot{u}}$, but
generally nonlinear in all other quantities (though often linear in the
external inputs $\bm{r}$).  In this formalism, both independent and dependent
quantities (coordinates and speeds) appear in these equations; we will address
this is shortly.  This following calculations formalism is exceedingly generic;
in practice there is a lot of structure in equations
(\ref{eq:nonholonomic}-\ref{eq:dyndiffs}) that can be leveraged to simplify the
following calculations.  This will be expanded upon below.

The system described by equations \ref{eq:holonomic}-\ref{eq:dyndiffs} has
$o-m$ velocity degrees of freedom.  In state space form, the trajectories of
the system exist in a $p \triangleq n - l + o - m$ dimensional space, though
conserved quantities (e.g., total mechanical energy) may restrict trajectories
to a lower dimensional manifold embedded in this $p$ dimensional space.  In
general this cannot be assumed, however, especially in the presence of external
inputs which may change the total mechanical energy of the system.  While more
than $p$ of the state variables may be of interest (i.e., dependent coordinates
or speeds may be of interest in themselves), only $p$ quantities are truly
independent.  Which of these $p$ coordinates and speeds should be chosen as
independent out of the possible $n + o$ may be obvious for small systems where
there is intuition about the system behavior, but in general the optimal choice
depends on how the configuration $\bm{q}$ affects the gradients of equations
\ref{eq:holonomic} and \ref{eq:nonholonomic}.  On the other hand, in systems
with cyclic (ignorable) coordinates, fewer than $p$ state variables may be of
interest; one example is the Whipple bicycle model where $p=10$ but 5 of these
state variables are cyclic coordinates (coordinates of rear wheel contact
point, heading, and both wheel angles) and are often not of much interest in
themselves, nor are they neccessary for most analyses.

We begin with a first order Taylor series expansion of equations
(\ref{eq:holonomic})-(\ref{eq:dyndiffs}) about an equilibrium
$\bm{q}=\bm{q}^*$, $\bm{\dot{q}}=\bm{\dot{q}}^*$, $\bm{u}=\bm{u}^*$,
$\bm{\dot{u}}=\bm{\dot{u}}^*$, $\bm{r}=\bm{r}^*$ that satisfies equations
(\ref{eq:holonomic})-(\ref{eq:dyndiffs}).  Determining these equilibrium values
generally involves solving (\ref{eq:holonomic})-(\ref{eq:dyndiffs}).  In the interest of brevity, we omit writing this equilibrium after
each gradient in the calculations below; all are evaluated at these equilibrium
conditions.  Taylor series expansion of the constraint equations
(\ref{eq:holonomic}) and (\ref{eq:nonholonomic}) about the equilibrium is
\begin{align}
  \label{eq:holonomic_expansion}
  \bm{f}_{h}(\bm{q}, t) &\approx \underbrace{f_{h}(\bm{q}^*, t)}_{\bm{0}} +
    \nabla_{\bm{q}}f_{h} \cdot \delta \bm{q}\\
  \label{eq:nonholonomic_expansion}
  \bm{f}_{nh}(\bm{q}, \bm{u}, t) &\approx \underbrace{\bm{f}_{nh}(\bm{q}^*,
  \bm{u}^*, t)}_{\bm{0}} +  \nabla_{\bm{q}}f_{nh} \cdot \delta \bm{q} +
  \nabla_{\bm{u}}f_{nh} \cdot \delta \bm{u}
\end{align}
and the first terms are identically zero because of the assumption that the
equilibrium point satisfies the constraints.  The Taylor series expansions of
the functions appearing in the left hand side of equations (\ref{eq:kindiffs})
are
\begin{align}
  \label{eq:f0_expansion}
  \bm{f}_{0}(\bm{q}, \bm{\dot{q}}, t) &\approx \bm{f}_{0}(\bm{q}^*,
  \bm{\dot{q}}^*, t) + \nabla_{\bm{q}}\bm{f}_{0} \cdot \delta\bm{q} +
  \nabla_{\bm{\dot{q}}}\bm{f}_{0} \cdot \delta\bm{\dot{q}}\\
  \label{eq:f1_expansion}
  \bm{f}_{1}(\bm{q}, \bm{u}, t) &\approx \bm{f}_{1}(\bm{q}^*,
  \bm{u}^*, t) + \nabla_{\bm{q}}\bm{f}_{1} \cdot \delta\bm{q} +
  \nabla_{\bm{u}}\bm{f}_{1} \cdot \delta\bm{u}
\end{align}
Summing (\ref{eq:f0_expansion}) and (\ref{eq:f1_expansion}) and recognizing
that the sum of the first term on the right hand side of each equation must
equal zero, we obtain
\begin{align}
  \label{eq:f0_plus_f1_expansion}
  \bm{f}_{0}(\bm{q}, \bm{\dot{q}}, t) + \bm{f}_{1}(\bm{q}, \bm{u}, t) &\approx
  \nabla_{\bm{q}}(\bm{f}_{0} + \bm{f}_{1}) \cdot \delta\bm{q} +
  \nabla_{\bm{\dot{q}}}\bm{f}_{0} \cdot \delta\bm{\dot{q}} +
  \nabla_{\bm{u}}\bm{f}_{1} \cdot \delta\bm{u}
\end{align}
Similarly, a Taylor series expansion of the functions appearing in
the left hand side of (\ref{eq:dyndiffs}) yields
\begin{align}
  \label{eq:f2_expansion}
  \bm{f}_{2}(\bm{q}, \bm{\dot{u}}, t) &\approx
      \bm{f}_{2}(\bm{q}^*, \bm{\dot{u}}^*, t) +
      \nabla_{\bm{q}}\bm{f}_{2} \cdot \delta\bm{q}
      + \nabla_{\bm{\dot{u}}}\bm{f}_{2} \cdot \delta\bm{\dot{u}}\\
  \bm{f}_{3}(\bm{q}, \bm{\dot{q}}, \bm{u}, \bm{r}, t) &\approx
  \bm{f}_{3}(\bm{q}^*, \bm{\dot{q}}^*, \bm{u}^*, \bm{r}^*, t) +
  \nabla_{\bm{q}}\bm{f}_{3} \cdot \delta\bm{q}\notag\\
  \label{eq:f3_expansion}
  &+ \nabla_{\bm{\dot{q}}}\bm{f}_{3} \cdot \delta\bm{\dot{q}}
  + \nabla_{\bm{u}}\bm{f}_{3} \cdot \delta \bm{u}
  + \nabla_{\bm{r}}\bm{f}_{3} \cdot \delta\bm{r}
\end{align}
Summing (\ref{eq:f2_expansion}) and (\ref{eq:f3_expansion}) and recognizing
that the sum of the first term on the right hand sides of these equations must
equal zero, we obtain
\begin{align}
  \bm{f}_{2}(\bm{q}, \bm{\dot{u}}, t) + \bm{f}_{3}(\bm{q}, \bm{\dot{q}},
  \bm{u}, \bm{r}, t) &\approx \nabla_{\bm{q}}(\bm{f}_2 + \bm{f}_3) \cdot
  \delta\bm{q} + \nabla_{\bm{\dot{q}}}\bm{f}_{3} \cdot \delta\bm{\dot{q}}\notag\\
  \label{eq:f2_plus_f3_expansion}
  &+ \nabla_{\bm{u}}\bm{f}_{3} \cdot \delta\bm{u} +
  \nabla_{\bm{\dot{u}}}\bm{f}_{2} \cdot \delta\bm{\dot{u}} + \nabla_{\bm{r}}\bm{f}_{3} \cdot \delta\bm{r}
\end{align}

Utilizing equations (\ref{eq:kindiffs}), (\ref{eq:f0_plus_f1_expansion}),
(\ref{eq:dyndiffs}), and (\ref{eq:f2_plus_f3_expansion}), we introduce the
following quantities
\begin{align}
  M_{00} &\triangleq \nabla_{\bm{\dot{q}}}\bm{f}_0 \\
  M_{11} &\triangleq \nabla_{\bm{\dot{u}}}\bm{f}_2 \\
  A_{00} &\triangleq -\nabla_{\bm{q}}(\bm{f}_0 + \bm{f}_1) \\
  A_{01} &\triangleq -\nabla_{\bm{u}}\bm{f}_1 \\
  A_{10} &\triangleq \nabla_{\bm{q}}(\bm{f}_2 + \bm{f}_3)
  -\nabla_{\bm{\dot{q}}}\bm{f}_3 (\nabla_{\bm{\dot{q}}}\bm{f}_{0})^{-1}
  \nabla_{\bm{q}}(\bm{f}_0 + \bm{f}_1) \\
  A_{11} &\triangleq \nabla_{\bm{u}} \bm{f}_{3} - \nabla_{\bm{\dot{q}}}\bm{f}_{3}
  (\nabla_{\bm{\dot{q}}}\bm{f}_0)^{-1} \nabla_{\bm{u}}\bm{f}_{1} \\
  B_{1} &\triangleq \nabla_{\bm{r}}\bm{f}_{3}
\end{align}

Equating the right hand side of equations (\ref{eq:f0_plus_f1_expansion}) and
(\ref{eq:f2_plus_f3_expansion}) each to zero, eliminating $\bm{\dot{q}}$ from
(\ref{eq:f2_plus_f3_expansion}), and rewriting in matrix form, we obtain
\begin{align}
  \label{eq:state_space_unconstrained}
  \left[
    \begin{array}{cc}
      M_{00} & \bm{0}_{n \times o} \\
      \bm{0}_{o \times n} & M_{11}
    \end{array}
    \right]
    \left[
      \begin{array}{c}
        \delta \bm{\dot{q}} \\
        \delta \bm{\dot{u}}
      \end{array}
    \right]
   &=
   \left[
     \begin{array}{cc}
       A_{00} & A_{01} \\
       A_{10} & A_{11}
     \end{array}
   \right]
    \left[
      \begin{array}{c}
        \delta \bm{q} \\
        \delta \bm{u}
      \end{array}
    \right]
    +
    \left[
      \begin{array}{c}
        \bm{0}_{n \times q} \\
        B_{1}
      \end{array}
    \right]
    \delta \bm{r}
\end{align}
Equation (\ref{eq:state_space_unconstrained}) has a state space of dimension $n
+ o$, yet  only $p = n - l + o - m$ of these quantities are independent, due to
the constraints (equations (\ref{eq:holonomic} and (\ref{eq:nonholonomic})).
To address this issue, one must pick a particular set of independent
coordinates and speeds and appeal to equations (\ref{eq:holonomic_expansion})
and (\ref{eq:nonholonomic_expansion}).  To this end, consider the following
partitioning of the generalized coordinates and generalized speeds:
\begin{equation*}
  \tilde{\bm{q}} \triangleq \left[\begin{array}{cc}\bm{q}_{i} &
      \bm{q}_{d}\end{array}\right]^{T} =  P_{q}^{-1} \bm{q}
      \qquad\qquad
  \tilde{\bm{u}} \triangleq \left[\begin{array}{cc}\bm{u}_{i} &
      \bm{u}_{d}\end{array}\right]^{T} =  P_{u}^{-1} \bm{u}
\end{equation*}
where $P_q \in \mathbf{R}^{n \times n}$ and $P_u \in \mathbf{R}^{o \times o}$
are invertible permutation matrices which map an ordering which has the
independent quantities ($\bm{q}_{i}\in\mathbf{R}^{n-l},\,
\bm{u}_{i}\in\mathbf{R}^{o-m})$ first, followed by the dependent quantities
($\bm{q}_{d}\in\mathbf{R}^{l},\, \bm{u}_{d}\in\mathbf{R}^{m}$) to the
original ordering of the coordinates and speeds.  We use the notation $P_{qi}$
and $P_{qd}$ to denote the first $n-l$ and last $l$ columns of $P_q$,
respectively; similarly, $P_{ui}$ is the first $o-m$ colums of $P_{u}$ while
$P_{ud}$ is the last $m$ columns of $P_u$.

Setting the right hand side of equation (\ref{eq:holonomic_expansion}) equal to
zero, and making use of $P_q$, we have
\begin{align}
  \nabla_{\bm{q}}\bm{f}_{h} P_{q} \cdot \delta \bm{\tilde{q}} &=
  \nabla_{\bm{q}}\bm{f}_{h} P_{qi} \cdot \delta \bm{q_i} +
  \nabla_{\bm{q}}\bm{f}_{h} P_{qd} \cdot \delta \bm{q_d}\notag\\
  &= \bm{0}\notag\\
  \implies \delta \bm{q}_d &= -(\nabla_{\bm{q}}\bm{f}_{h} P_{qd})^{-1}
  (\nabla_{\bm{q}}\bm{f}_{h} P_{qi}) \cdot \delta \bm{q}_i \notag\\
  \implies \delta \bm{q} &= \left[ I_{n \times n} - P_{qd}(\nabla_{\bm{q}}
    \bm{f}_{h} P_{qd})^{-1} \nabla_{\bm{q}} \bm{f}_{h} \right] P_{qi} \cdot \delta
    \bm{q}_i
\end{align}
For convenience, we define
\begin{equation}
  \label{eq:C_0}
  C_0 \triangleq \left[ I_{n \times n} - P_{qd}(\nabla_{\bm{q}}
    \bm{f}_{h} P_{qd})^{-1} \nabla_{\bm{q}} \bm{f}_{h} \right] P_{qi}
\end{equation}
which enables us to write
\begin{align}
  \label{eq:delta_q}
  \delta \bm{q} &= C_0 \cdot \delta \bm{q}
\end{align}

Applying the same approach to (\ref{eq:nonholonomic_expansion}), we obtain
\begin{align}
\nabla_{\bm{q}}f_{nh} \cdot \delta \bm{q} +
\nabla_{\bm{u}}f_{nh} \cdot \delta \bm{u} &= \nabla_{\bm{q}} \bm{f}_{nh} \cdot
\delta \bm{q} + \nabla_{\bm{u}} \bm{f}_{nh} P_{ui} \cdot \delta \bm{u}_i +
\nabla_{\bm{u}} \bm{f}_{nh} P_{ud} \cdot \delta \bm{u}_d \notag\\
&= \bm{0}\notag\\
\implies \delta \bm{u}_d &= -\left(\nabla_{\bm{u}} \bm{f}_{nh}
P_{ud}\right)^{-1}\left[\nabla_{\bm{q}}\bm{f}_{nh} \cdot \delta\bm{q} +
  \nabla_{\bm{u}} \bm{f}_{nh} P_{ui} \cdot \delta \bm{u}_i \right]\notag
\end{align}
We define
\begin{align}
  \label{eq:C_1}
  C_1 &\triangleq -\left(\nabla_{\bm{u}} \bm{f}_{nh} P_{ud}\right)^{-1}
  \nabla_{\bm{q}}\bm{f}_{nh} \\
  \label{eq:C_2}
  C_2 &\triangleq -\left(\nabla_{\bm{u}} \bm{f}_{nh} P_{ud}\right)^{-1} \nabla_{\bm{u}} \bm{f}_{nh} P_{ui}
\end{align}
which enables us to write
\begin{align}
  \label{eq:delta_u}
  \delta \bm{u} &= C_1 \cdot \delta \bm{q} + C_2 \cdot \delta \bm{u}_i\notag\\
  &= C_1 C_0 \cdot \delta \bm{q}_i + C_2 \cdot \delta \bm{u}_i
\end{align}

>>>>

The complete configuration of the system is assumed to be described by
generalized coordinates $\bm{q}\in\bm{R}^n$ and its velocity completely
described by generalized speeds $\bm{u}\in\bm{R}^o$, with $n \ne o$ in general.
All coordinates and speeds, not only the independent ones, are represented by
$\bm{q}$ and $\bm{u}$.  The system is assumed to be governed by the
relationships described in table \ref{table:assumptions}.  The system is
described by $l + m + n + o$ equations which relate $2n + 2o + 2p + 1$
quantities, which implies that $n - l + o - m + p + 1$ of these quantities are
independent.  This paper discusses the first order relationship between these
independent quantities ($n-l$ independent coordinates, $o-m$ independent
speeds, $p$ control inputs, and time $t$) and the time derivatives of the $n$
coordinates and $o$ speeds.
%It is assumed that there are $l$
%configuration constraints (), and $m$ velocity constraints , and $m$ acceleration
%constraints (). Two functions $ relate the coordinate time derivatives to the generalized speeds,
%and two more functions $\bm{f}_2 : \mathbf{R}^n \times \mathbf{R}^o \times
%\mathbf{R} \to \mathbf{R}^{o-m}$ and $\bm{f}_3 : \mathbf{R}^n
%\times \mathbf{R}^n \times \mathbf{R}^o \times \mathbf{R}^p \times \mathbf{R}
%\to \mathbf{R}^{o-m}$ relate the time derivatives of the speeds $\bm{\dot{u}}$ to
%the externally applied inputs $\bm{r}\in\bm{R}^q$, coordinates $\bm{q}$ and
%their time derivatives $\bm{\dot{q}}$,  speeds $\bm{u}$, and time $t$. These
%assumptions can be summarized by the following equations
%\begin{align}
%  \label{eq:configuration}
%  \bm{f}_{c}(\bm{q}, t) &= \bm{0}\\
%  \label{eq:velocity}
%  \bm{f}_{v}(\bm{q}, \bm{u}, t) &= \bm{0}\\
%  \label{eq:acceleration}
%  \bm{f}_{a}(\bm{q}, \bm{\dot{q}}, \bm{u}, \bm{\dot{u}}, t) &= \bm{0}\\
%  \label{eq:kindiffs}
%  \bm{f}_{0}(\bm{q}, \bm{\dot{q}}, t)
%    + \bm{f}_{1}(\bm{q}, \bm{u}, t) &= \bm{0} \\
%  \label{eq:dyndiffs}
%  \bm{f}_{2}(\bm{q}, \bm{\dot{u}}, t)
%    + \bm{f}_{3}(\bm{q}, \bm{\dot{q}}, \bm{u}, \bm{r}, t) & = \bm{0}
%\end{align}
\begin{table}[htbp]
  \centering
  \caption{Constrained multibody system governing equations}
  \begin{tabular}[c]{r @{ $=$ } l l}
    \multicolumn{2}{c}{Equation} & Mathematical definition \\
    \hline
    $\bm{f}_{c}(\bm{q}, t)$ & $\bm{0}$ & $f_c : \mathbf{R}^n \times \mathbf{R} \to
\mathbf{R}^l$ \\
$\bm{f}_{v}(\bm{q}, \bm{u}, t)$ & $\bm{0}$ & $f_v : \mathbf{R}^n \times
\mathbf{R}^o \times \mathbf{R} \to \mathbf{R}^m$ \\
$\bm{f}_{a}(\bm{q}, \bm{\dot{q}}, \bm{u}, \bm{\dot{u}}, t)$ & $\bm{0}$ &
$f_a : \mathbf{R}^n \times \mathbf{R}^n \times \mathbf{R}^o \times
\mathbf{R}^o \times \mathbf{R} \to \mathbf{R}^m$ \\
$\bm{f}_{0}(\bm{q}, \bm{\dot{q}}, t) + \bm{f}_{1}(\bm{q}, \bm{u}, t)$ &
$\bm{0}$ & $\bm{f}_0 : \mathbf{R}^n \times \mathbf{R}^n \times \mathbf{R} \to
\mathbf{R}^n$\\
\multicolumn{2}{c}{} & $\bm{f}_1 : \mathbf{R}^n \times \mathbf{R}^o \times \mathbf{R}
\to \mathbf{R}^n$ \\
$\bm{f}_{2}(\bm{q}, \bm{\dot{u}}, t) + \bm{f}_{3}(\bm{q}, \bm{\dot{q}}, \bm{u}, \bm{r}, t)$ &
$\bm{0}$ & $\bm{f}_2 : \mathbf{R}^n \times \mathbf{R}^o \times \mathbf{R} \to
\mathbf{R}^{o-m}$,\\
\multicolumn{2}{c}{} & $\bm{f}_3 : \mathbf{R}^n \times \mathbf{R}^n \times
\mathbf{R}^o \times \mathbf{R}^p \times \mathbf{R} \to \mathbf{R}^{o-m}$
  \end{tabular}
  \label{table:assumptions}
\end{table}
The first three equations in table \ref{table:assumptions} represent
constraints derived purely from kinematic considerations at the configuration,
velocity, and acceleration levels, respectively.  The velocity constraints may
be nonholonomic or differentiated holonomic, while the acceleration constraints
are assumed to be differentiated versions of the velocity constraints.  The
fourth equation represents kinematic differential equations which relate
$\bm{\dot{q}}$ and $\bm{u}$ and will be linear in both of these quantities.
The fifth equation represents the constraint free nonholonomic dynamic
differential equations~\cite{Kane1985}, and will necessarily by linear in
$\bm{\dot{u}}$ and input forces and torques $\bm{r}$, but generally nonlinear
in all other quantities.  This formalism is exceedingly generic; in practice,
the structure of the functions in table \ref{table:assumptions} can be
leveraged to simplify the following calculations.  This will be discussed
subsequently.

%The system described by equations \ref{eq:configuration}-\ref{eq:dyndiffs} has
%$o-m$ velocity degrees of freedom.
Trajectories of the system exist on a $p \triangleq n - l + o - m$ dimensional
manifold embedded in a $n + o$ dimensional space, though conserved quantities
(e.g., total mechanical energy) may restrict trajectories to a lower
dimensional manifold.  In general this cannot be assumed, however, especially
in the presence of external control inputs $\bm{r}$ which may change the total
mechanical energy of the system.  While more than $p$ of the independent state
variables may be of interest (i.e., dependent coordinates or speeds may be of
interest in themselves), only $p$ quantities are truly independent.  Which of
these $p$ coordinates and speeds should be chosen as independent out of the
possible $n + o$ may be obvious for small systems where there is intuition
about the system behavior, but in general the optimal choice depends on how the
configuration $\bm{q}$ affects the gradients of $f_c$ and $f_v$.  On the other
hand, in systems with cyclic/ignorable coordinates (i.e., coordinates which do
not appear in the dynamic equations of motion), fewer than $p$ state
variables may be of interest; one example is the Whipple bicycle model where
$p=10$ but 5 of these state variables are cyclic coordinates (coordinates of
rear wheel contact point, heading, and both wheel angles) and are often not of
much interest in themselves, nor are they necessary for most analyses.

We begin with a first order Taylor series expansion of the equations in Table
\ref{table:assumptions} about $\bm{q}=\bm{q}^*$, $\bm{\dot{q}}=\bm{\dot{q}}^*$,
$\bm{u}=\bm{u}^*$, $\bm{\dot{u}}=\bm{\dot{u}}^*$, $\bm{r}=\bm{r}^*$; it is
assumed that all of the equations in Table \ref{table:assumptions} are
satisfied by these quantities.  In the interest of brevity, we omit writing
this equilibrium after each gradient in the calculations below; all are
evaluated at these equilibrium conditions.  Expansion of the three constraint
equations yields
\begin{align}
  \label{eq:configuration_expansion}
  \bm{f}_{c}(\bm{q}, t) &\approx \underbrace{f_{c}(\bm{q}^*, t)}_{\bm{0}} +
    \nabla_{\bm{q}}f_{c} \cdot \delta \bm{q}\\
  \label{eq:velocity_expansion}
  \bm{f}_{v}(\bm{q}, \bm{u}, t) &\approx \underbrace{\bm{f}_{v}(\bm{q}^*,
  \bm{u}^*, t)}_{\bm{0}} +  \nabla_{\bm{q}}f_{v} \cdot \delta \bm{q} +
  \nabla_{\bm{u}}f_{v} \cdot \delta \bm{u} \\
  \label{eq:acceleration_expansion}
  \bm{f}_{a}(\bm{q}, \bm{\dot{q}}, \bm{u}, \bm{\dot{u}}, t) &\approx
  \underbrace{\bm{f}_{a}(\bm{q}^*, \bm{\dot{q}}^*, \bm{u}^*, \bm{\dot{u}}^*,
t)}_{\bm{0}} +  \nabla_{\bm{q}}f_{a} \cdot \delta \bm{q} + \nabla_{\bm{\dot{q}}}f_{a}
\cdot \delta \bm{\dot{q}} \notag\\
&+ \nabla_{\bm{u}}f_{a} \cdot \delta \bm{u} + \nabla_{\bm{\dot{u}}}f_{a} \cdot
\delta \bm{\dot{u}}
\end{align}
The first terms are identically zero because of the assumption that the
equilibrium point satisfies the constraints.  The Taylor series expansion of
the kinematic differential equations is
\begin{align}
  \label{eq:f0_expansion}
  \bm{f}_{0}(\bm{q}, \bm{\dot{q}}, t) &\approx \bm{f}_{0}(\bm{q}^*,
  \bm{\dot{q}}^*, t) + \nabla_{\bm{q}}\bm{f}_{0} \cdot \delta\bm{q} +
  \nabla_{\bm{\dot{q}}}\bm{f}_{0} \cdot \delta\bm{\dot{q}}\\
  \label{eq:f1_expansion}
  \bm{f}_{1}(\bm{q}, \bm{u}, t) &\approx \bm{f}_{1}(\bm{q}^*,
  \bm{u}^*, t) + \nabla_{\bm{q}}\bm{f}_{1} \cdot \delta\bm{q} +
  \nabla_{\bm{u}}\bm{f}_{1} \cdot \delta\bm{u}
\end{align}
Summing (\ref{eq:f0_expansion}) and (\ref{eq:f1_expansion}) and recognizing
that the sum of the first term on the right hand side of each equation must
equal zero, we obtain
\begin{align}
  \label{eq:f0_plus_f1_expansion}
  \bm{f}_{0}(\bm{q}, \bm{\dot{q}}, t) + \bm{f}_{1}(\bm{q}, \bm{u}, t) &\approx
  \nabla_{\bm{q}}(\bm{f}_{0} + \bm{f}_{1}) \cdot \delta\bm{q} +
  \nabla_{\bm{\dot{q}}}\bm{f}_{0} \cdot \delta\bm{\dot{q}} +
  \nabla_{\bm{u}}\bm{f}_{1} \cdot \delta\bm{u}
\end{align}
Similarly, a Taylor series expansion of the dynamic differential equations, we
obtain
\begin{align}
  \label{eq:f2_expansion}
  \bm{f}_{2}(\bm{q}, \bm{\dot{u}}, t) &\approx
      \bm{f}_{2}(\bm{q}^*, \bm{\dot{u}}^*, t) +
      \nabla_{\bm{q}}\bm{f}_{2} \cdot \delta\bm{q}
      + \nabla_{\bm{\dot{u}}}\bm{f}_{2} \cdot \delta\bm{\dot{u}}\\
  \bm{f}_{3}(\bm{q}, \bm{\dot{q}}, \bm{u}, \bm{r}, t) &\approx
  \bm{f}_{3}(\bm{q}^*, \bm{\dot{q}}^*, \bm{u}^*, \bm{r}^*, t) +
  \nabla_{\bm{q}}\bm{f}_{3} \cdot \delta\bm{q}\notag\\
  \label{eq:f3_expansion}
  &+ \nabla_{\bm{\dot{q}}}\bm{f}_{3} \cdot \delta\bm{\dot{q}}
  + \nabla_{\bm{u}}\bm{f}_{3} \cdot \delta \bm{u}
  + \nabla_{\bm{r}}\bm{f}_{3} \cdot \delta\bm{r}
\end{align}
Summing (\ref{eq:f2_expansion}) and (\ref{eq:f3_expansion}) and recognizing
that the sum of the first term on the right hand sides of these equations must
equal zero, we obtain
\begin{align}
  \bm{f}_{2}(\bm{q}, \bm{\dot{u}}, t) + \bm{f}_{3}(\bm{q}, \bm{\dot{q}},
  \bm{u}, \bm{r}, t) &\approx \nabla_{\bm{q}}(\bm{f}_2 + \bm{f}_3) \cdot
  \delta\bm{q} + \nabla_{\bm{\dot{q}}}\bm{f}_{3} \cdot \delta\bm{\dot{q}}\notag\\
  \label{eq:f2_plus_f3_expansion}
  &+ \nabla_{\bm{u}}\bm{f}_{3} \cdot \delta\bm{u} +
  \nabla_{\bm{\dot{u}}}\bm{f}_{2} \cdot \delta\bm{\dot{u}} + \nabla_{\bm{r}}\bm{f}_{3} \cdot \delta\bm{r}
\end{align}

Equating the right hand sides of equations (\ref{eq:f0_plus_f1_expansion}),
(\ref{eq:acceleration_expansion}),
and (\ref{eq:f2_plus_f3_expansion}) to zero (as per table
\ref{table:assumptions}), and introducing the following definitions
\begin{align}
\tilde{M}_{qq}  &\triangleq \nabla_{\bm{\dot{q}}}\bm{f}_0 \\
\tilde{M}_{uuc} &\triangleq \nabla_{\bm{\dot{u}}}\bm{f}_a \\
\tilde{M}_{uud} &\triangleq \nabla_{\bm{\dot{u}}}\bm{f}_2 \\
\tilde{A}_{qq}  &\triangleq -\nabla_{\bm{q}}(\bm{f}_0 + \bm{f}_1) \\
\tilde{A}_{qu}  &\triangleq -\nabla_{\bm{u}}\bm{f}_1 \\
\tilde{A}_{uqc} &\triangleq -\left[ \nabla_{\bm{\dot{q}}}\bm{f}_a \tilde{M}_{qq}^{-1} \tilde{A}_{qq} + \nabla_{\bm{q}} \bm{f}_a\right] \\
\tilde{A}_{uuc} &\triangleq -\left[ \nabla_{\bm{\dot{q}}}\bm{f}_a \tilde{M}_{qq}^{-1} \tilde{A}_{qu} + \nabla_{\bm{u}} \bm{f}_a\right] \\
\tilde{A}_{uqd} &\triangleq -\left[ \nabla_{\bm{\dot{q}}}\bm{f}_3 \tilde{M}_{qq}^{-1} \tilde{A}_{qq} + \nabla_{\bm{q}} (\bm{f}_2 + \bm{f}_3)\right] \\
\tilde{A}_{uud} &\triangleq -\left[ \nabla_{\bm{\dot{q}}}\bm{f}_3 \tilde{M}_{qq}^{-1} \tilde{A}_{qu} + \nabla_{\bm{u}} \bm{f}_u\right] \\
\tilde{B}_{u}   &\triangleq \nabla_{\bm{r}}\bm{f}_{3}
\end{align}
enables the unconstrained linear state space equations to be written as
\begin{align}
  \label{eq:state_space_unconstrained}
  \left[
    \begin{array}{cc}
      \tilde{M}_{qq} & \bm{0}_{n \times o} \\
      \bm{0}_{m \times n} & \tilde{M}_{uuc} \\
      \bm{0}_{(o-m) \times n} & \tilde{M}_{uud}
    \end{array}
    \right]
    \left[
      \begin{array}{c}
        \delta \bm{\dot{q}} \\
        \delta \bm{\dot{u}}
      \end{array}
    \right]
   &=
   \left[
     \begin{array}{cc}
       \tilde{A}_{qq} & \tilde{A}_{qu} \\
       \tilde{A}_{uqc} & \tilde{A}_{uuc} \\
       \tilde{A}_{uqd} & \tilde{A}_{uud}
     \end{array}
   \right]
    \left[
      \begin{array}{c}
        \delta \bm{q} \\
        \delta \bm{u}
      \end{array}
    \right]
    +
    \left[
      \begin{array}{c}
        \bm{0}_{(n + m) \times q} \\
        \tilde{B}_{u}
      \end{array}
    \right]
    \delta \bm{r}
\end{align}
Equation (\ref{eq:state_space_unconstrained}) has a state space of dimension $n
+ o$, yet only $p = n - l + o - m$ of these quantities are independent.
To address this issue, a particular set of independent
coordinates and speeds must be selected. To this end, consider the following
partitioning of the generalized coordinates and generalized speeds:
\begin{equation*}
  \tilde{\bm{q}} \triangleq \left[\begin{array}{cc}\bm{q}_{i} &
      \bm{q}_{d}\end{array}\right]^{T} =  P_{q}^{-1} \bm{q}
      \qquad\qquad
  \tilde{\bm{u}} \triangleq \left[\begin{array}{cc}\bm{u}_{i} &
      \bm{u}_{d}\end{array}\right]^{T} =  P_{u}^{-1} \bm{u}
\end{equation*}
where $P_q \in \mathbf{R}^{n \times n}$ and $P_u \in \mathbf{R}^{o \times o}$
are invertible permutation matrices which map an ordering which has the
independent quantities ($\bm{q}_{i}\in\mathbf{R}^{n-l},\,
\bm{u}_{i}\in\mathbf{R}^{o-m})$ first, followed by the dependent quantities
($\bm{q}_{d}\in\mathbf{R}^{l},\, \bm{u}_{d}\in\mathbf{R}^{m}$) to the
original ordering of the coordinates and speeds.  We use the notation $P_{qi}$
and $P_{qd}$ to denote the first $n-l$ and last $l$ columns of $P_q$,
respectively; similarly, $P_{ui}$ is the first $o-m$ colums of $P_{u}$ while
$P_{ud}$ is the last $m$ columns of $P_u$.
